Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-25 paper.tex
Partial update of the LS section

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs, and this representation should be well-suited to storing
recombinant viral genealogies, even those with millions of sampled sequence,
such as SARS-CoV-2 (with currently over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the ``tskit'' software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

@@

\subsection{The Li and Stephens model}
[Edit down, cutting the text in the fig caption significantly]

The Li and Stephens (LS) model is
a convenient probabilistic approximation to the coalescent with recombination
\citep{Hudson1983-properties}, itself a generative model of a genealogical process with
recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
in which a focal genome is a sequence of nucleotides (observed states) that are
probabilistically emitted as an imperfect mosaic of a set of reference genomes
(hidden states). As such, an efficient HMM machinery can be employed to make
statements about parameters of interest and highlight potential sources of
error in genotyping and sequencing data. The LS model is used in a wide variety
of applications in genomics. For example, it is the engine that drives
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and
it can be used to estimate mutation rates (e.g., REF), recombination rates
\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.

\begin{figure} \centering
\includegraphics[width=0.6\textwidth]{figures/ls_diagram_covid.pdf}
% FIXME this caption is too long
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS) model.
The LS model in its implementation as an HMM is governed by the transition and
emission matrices, $Q$ and $E$, respectively. Each may be a function of the
reference panel members, but transitions are generally assumed to be
independent of the hidden states themselves (shown in the pink panel). This
dramatically increases performance as the state space drops to two states
(i.e., switching or not switching). The recombination probability at site
$\ell$ is $r\ell$, while n denotes the number of members in the reference
panel. Emissions may also be a function of the nucleotide states, but in our
case, we assume that all observed alleles (in both the reference panel and
focal sequence, $a\ell$) at the site under consideration are equally likely and
occur with mutation probability $\ell$ (shown in the blue panel). These two
processes, transition and emission (which encode recombination and mutation,
respectively), define the generative process of the HMM, by which an imperfect
mosaic is emitted from a reference panel of samples and nodes from the ARG
inserted at earlier collection dates (shown in the lower panel). Here, the
unseen states in the reference panel are shown as coloured lines enclosed by
the grey box, and the true path through the data which leads to the emitted
focal sequence is displayed by the black arrow. Examples of transition and
emission probabilities along this trajectory are shown by the red and blue
arrows, respectively. Black crosses along the focal sequence show genotyping
errors or mutations. Note that in practice, we do not know the pattern of
colours along the focal sequence and must infer it using the Viterbi
algorithm.} \end{figure}

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as genotyping or sequencing error). Then, we use the Viterbi algorithm
\citep{Viterbi1967-ol} to find the most likely copying path through the
sequences, given a set of reference sequences, $Q$, and $E$.

Because the HMM is memoryless, the potential parental lineages are
independently assessed at each recombination breakpoint. This can result in the
inference of a recombinant lineage descending immediately from three (or even
more) parental lineages. We find that often two of the three parents are
identical or similar in sequence in the segments contributed to the recombinant
lineage. A more parsimonious explanation would be to switch back to the
original parent sequence. In the future, a pragmatic solution would be to
detect such cases during the ARG inference stage and to adjust the copying
paths accordingly (see ``Improvements'', below). Meanwhile, however, we urge
caution when interpreting the parentage of lineages with multiple breakpoints.

@@

\begin{table} \centering
\begin{tabular}{ll|clll}
\toprule
A Group (XA) & Jackson        & 21,255–21,613 & B.1.177/B.1.1.7 \\
             & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
\midrule
B Group & Jackson        &  6,528--6,954 & B.1.36/B.1.1.7  \\
        & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
\midrule
C Group & Jackson        &  25,997--27,972 &  B.1.1.7/B.1.221 \\
        & \texttt{sc2ts} &  25,996--27,441 &  B.1.1.7/B.1.221 \\
\midrule
D Group & Jackson        &  21,575--23,063 &  B.1.36.17/B.1.1.7 \\
        & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
\midrule
CAMC-CBA018 & Jackson        &  20,389--21,254 & B.1.177/B.1.1.7 \\
            & \texttt{sc2ts} & 16,177--21,255 & B.1.177/B.1.1.7 \\
\midrule
\end{tabular}
\caption{\label{tab:jackson}Comparison of recombination breakpoint intervals and parent lineages for a subset of the data reported by \cite{Jackson2021-ik}. [FIXME ADD MORE DETAIL] More detailed results are given in Table~\ref{tab:jackson_supplement}.}
\end{table}

>>>>

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs, and this representation should be well-suited to storing
recombinant viral genealogies, even those with millions of sampled sequence,
such as SARS-CoV-2 (with currently over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the ``tskit'' software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

@@

\subsection{The Li and Stephens model}
The Li and Stephens (LS) model\citep{Li2003-ib}  is an approximation of the
coalescent with recombination~\citep{Hudson1983-properties}, which captures
many of the important features of the joint processes of mutation and
recombination. It is a Hidden Markov Model (HMM) in which a focal genome is a
sequence of nucleotides that are probabilistically emitted as
an imperfect mosaic of a set of reference genomes
(Figure~\ref{fig:ls_diagram}).
The LS model is used in a wide variety
of applications in genomics, including modern approaches to
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
and estimation of parameters such as
recombination rates \citep[e.g.][]{Hinch2011-tz},
and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.
See~\cite{Mcvean2019-linkage} for further review and discussion
of the LS model.

\begin{figure} \centering
\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
% FIXME this caption is too long
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
model, in which a query sequence (bottom) is described as an
imperfect mosaic of the sequences in a reference panel.
Black crosses along the focal sequence show genotyping
errors or mutations.
In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
the mutation probability is $\mu_\ell$ and $n$
denotes the size of the reference panel.
The Viterbi algorithm can be used to find a
``copying path'' through the reference panel for a given query sequence that
maximises the likelihood under these parameters.
% The LS HMM is governed by the transition and
% emission matrices, $Q$ and $E$, respectively.
% Each may be a function of the
% reference panel members, but transitions are generally assumed to be
% independent of the hidden states themselves (shown in the pink panel). This
% dramatically increases performance as the state space drops to two states
% (i.e., switching or not switching).
% Emissions may also be a function of the nucleotide states, but in our
% case, we assume that all observed alleles (in both the reference panel and
% focal sequence, $a\ell$) at the site under consideration are equally likely and
% occur with mutation probability $\ell$ (shown in the blue panel).
% These two
% processes, transition and emission (which encode recombination and mutation,
% respectively), define the generative process of the HMM, by which an imperfect
% mosaic is emitted from a reference panel.
 % of samples and nodes from the ARG
% inserted at earlier collection dates (shown in the lower panel). Here, the
% unseen states in the reference panel are shown as coloured lines enclosed by
% the grey box, and the true path through the data which leads to the emitted
% focal sequence is displayed by the black arrow. Examples of transition and
% emission probabilities along this trajectory are shown by the red and blue
% arrows, respectively.
% Note that in practice, we do not know the pattern of
% colours along the focal sequence and must infer it using the Viterbi algorithm.
}
\end{figure}

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as genotyping or sequencing error). Then, we use the Viterbi algorithm
\citep{Viterbi1967-ol} to find the most likely copying path through the
sequences, given a set of reference sequences, $Q$, and $E$.
The relative rates of mutation and recombination are usually controlled by
the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
For this work we used a simpler formulation, which uses a single parameter
(the mismatch ratio, MMR) to control the relative importance of mutation
and recombination in the HMM. Specifically, an MMR value of $k$ will
prefer $k$ mismatches (mutations) to a recombination.[DUNCAN MORE DETAILS
HERE PLEASE]

[JK: TODO, expand this next paragraph to two, explaining a bit more that
we're searching among every node in the ARG, using the efficient tree
implementation in tsinfer, and a few words about how we can use the
HMM more, to do things like stochastic traceback, etc.]

@@

\begin{table} \centering
\begin{tabular}{ll|clll}
\toprule
A Group (XA) & Jackson        &  21,256–21,613 & B.1.177/B.1.1.7 \\
             & \texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
\midrule
B Group & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
        & \texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
\midrule
C Group & Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
        & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
\midrule
D Group & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
        & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
\midrule
CAMC-CBA018 & Jackson        &  20,390--21,254 & B.1.177/B.1.1.7 \\
            & \texttt{sc2ts} &  16,177--21,255 & B.1.177/B.1.1.7 \\
\midrule
\end{tabular}
\caption{\label{tab:jackson}Comparison of recombination breakpoint intervals and parent lineages for a subset of the data reported by \cite{Jackson2021-ik}. [FIXME ADD MORE DETAIL] More detailed results are given in Table~\ref{tab:jackson_supplement}.}
\end{table}
